---
title: "Low Inflation in an Island of High Prices. The Factors behind Price Stability in Switzerland"
format: html
editor: visual
draft: true
bibliography: inflation_switzerland.bib
---

## Introduction

With the recurrence of large-scale global supply shocks such as the COVID-19 crisis, the war in Ukraine, and more recently the closure of the Strait of Hormuz, inflation has made a strong return in developed economies. In recent years, the latter have experienced inflation levels not seen since the oil shocks of the 1970s. According to the KOF Economic Institute, inflation reached 8% in the United States and 8.4% in the Eurozone in 2022, before declining to 5.5% and 4.1%, respectively, in 2023. Inflation has not decreased below the target rate of 2% ever since. In Switzerland, inflation rose up to barely 2,8% in 2022 and 2,1% in 2023. Since 2024, inflation is already back to the Swiss National Bank (SNB) target range (0-2%) [@sturm_konjunkturanalyse_2023]. Given the challenge posed for price stability due to the global geopolitical context, the Swiss economy has indeed proved to be very resistant to inflationary pressures.  At the peak of the 2022-2023 global food crisis, food inflation reached almost 20% in March 2023 in the European Union while it reached a maximum of 6,8% in Switzerland (see Food Price Monitoring Tool by Eurostat). Swiss prices are among the most stable in the world since the last two decades, with an average inflation rate of its national consumer price index of only 0,6% between 2000 and 2025. In international comparison, only Japan has experienced lower inflation (0,5%), while average inflation rates usually ranged from 1,5 to 2,5% in developed economies (for instance 2% for the Eurozone and 2,5% for the United States). At the same time, Switzerland has one of the highest price level in the world. Its GDP price index is approximately 60% higher than the European Union (EU) average and final household expenditure prices are higher by 82% [@albouy_swiss_2026]. Switzerland is thus more known as a "high-price island" than for its resistance to inflation and Swiss policy makers are far more preocuppied by the country's high price level than by its rate of increase. 

What factors lie behind the high price level and relatively low inflation rates in Switzerland ? In the context of overlapping emergencies which make supply shocks more likely to multiply in the future, understanding how and why Switzerland is relatively more resistant to cost-push inflation can shed light on the current debates about the optimal economic policies for economic stabilization. Since monetary policy proved to be powerless against the resurgence of inflation, some economists advocated for less conventional policies such as price controls. The Swiss case is thus an interesting case study to explore what policies and institutions are effective against inflation.

We first review the existing literature on Swiss inflation and price stability, as well as the main economic developments of Switzerland which are relevant to the country's price dynamics. We then turn to a model, which remains undetermined at the present time as Michalis and Guillaume let me do all the work.

## Literature Review


```{r inflation capture by profit and labor}

library(tidyverse)
library(readxl)
library(lubridate)

inflation_income_decomposition <- read_excel("inflation_income_decomposition.xlsx", 
    sheet = "results")

inflation_income_decomposition <- 
  inflation_income_decomposition |>
  mutate(period = yq(period))


inflation_income_decomposition |>
  pivot_longer(c(Labor, Profit), names_to = "component", values_to = "value") |>
  ggplot(aes(x = period, y = value, fill = component)) +
  geom_col(position = "stack", width = 70) +
  geom_line(
    data = inflation_income_decomposition,
    aes(x = period, y = Inflation, color = "Inflation"),
    linewidth = 0.8, inherit.aes = FALSE
  ) +
  scale_fill_manual(
    values = c("Labor" = "#2980b9", "Profit" = "#e07b39"),
    name = NULL
  ) +
  scale_color_manual(values = c("Inflation" = "#c0392b"), name = NULL) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

```














```{r effective exchange rate snb}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(lubridate)

effective_exch_rate_snb <- read_excel("effective_exch_rate_snb.xlsx", 
    skip = 15)


effective_exch_rate_snb$period <-  ym(effective_exch_rate_snb$period)

effective_exch_rate_snb |> 
  ggplot(aes(x = period, y = `Overall index nominal`, color = "Overall index nominal"))+
  geom_line()+
  geom_line(aes(x = period, y = `Overall index real`, color = "Overall index real"), linetype = 2)+
  theme_minimal()+
  labs(
    title = "Nominal and real effective CHF exchange rate index",
    subtitle = "Monthly basis; December 2000 = 100",
    y = "Index point", x = ""
  )+
  theme(
    legend.title = element_blank(),
    legend.position = c(0.2, 0.8)
  )

```








```{r taux directeur}
#| echo: false
#| warning: false
#| message: false

library(rdbnomics)
library(tidyverse)
library(patchwork)

# --- Import données ---
hicp_ch_ea <- rdb("Eurostat", "PRC_HICP_MIDX",
                  dimensions = list(coicop = "CP00", unit = "I05", geo = c("CH", "EA")))

rates_raw <- rdb("BIS", "WS_CBPOL", dimensions = list(REF_AREA = c("CH", "XM")))

# --- Mise en forme taux directeurs (mensuel) ---
rates <- rates_raw |>
  filter(str_detect(series_name, "Monthly")) |>
  select(country = REF_AREA, period, value) |>
  mutate(
    country = case_when(
      country == "CH" ~ "BNS (Suisse)",
      country == "XM" ~ "BCE (Zone euro)"
    ),
    period = as.Date(period)
  ) |>
  filter(period >= as.Date("2005-01-01"))

# --- Inflation glissement annuel ---
hicp_yoy <- hicp_ch_ea |>
  select(geo, period, value) |>
  mutate(
    country = if_else(geo == "CH", "Suisse", "Zone euro"),
    yoy     = (value / lag(value, 12) - 1) * 100,
    .by = geo
  ) |>
  filter(!is.na(yoy), period >= as.Date("2005-01-01"))

# --- Zone grisée 2008-2015 ---
shade <- data.frame(
  xmin = as.Date("2008-09-01"), xmax = as.Date("2015-01-01"),
  ymin = -Inf, ymax = Inf
)

# --- Graphique 1 : inflation ---
p1 <- ggplot(hicp_yoy, aes(x = period, y = yoy, colour = country)) +
  geom_rect(data = shade, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "grey85", alpha = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c("Suisse" = "#c0392b", "Zone euro" = "#2980b9")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Inflation (HICP, glissement annuel)", y = "%", x = NULL, colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

# --- Graphique 2 : taux directeurs ---
p2 <- ggplot(rates, aes(x = period, y = value, colour = country)) +
  geom_rect(data = shade, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "grey85", alpha = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c("BNS (Suisse)" = "#c0392b", "BCE (Zone euro)" = "#2980b9")) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  labs(title = "Taux directeurs (BNS vs BCE)", y = "%", x = NULL, colour = NULL,
       caption = "Zone grisée : 2008–2015   |   Source : Eurostat, BIS via DBnomics") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p2_inset <- p2 + labs(caption = NULL, title = "Taux directeurs") +
  theme(legend.key.size = unit(0.4, "cm"),
        legend.text = element_text(size = 7),
        plot.title = element_text(size = 9),
        axis.text = element_text(size = 7))

p1 + inset_element(p2_inset, left = 0, bottom = 0.52, right = 0.44, top = 1)
```



```{r inflation ameco cpi}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(rdbnomics)

codes_variables <- c("BEL.3.0.0.0.ZCPIN", "CHE.3.0.0.0.ZCPIN", "DEU.3.0.0.0.ZCPIN", "DNK.3.0.0.0.ZCPIN", "FRA.3.0.0.0.ZCPIN", "GBR.3.0.0.0.ZCPIN", "ITA.3.0.0.0.ZCPIN", "LUX.3.0.0.0.ZCPIN", "USA.3.0.0.0.ZCPIN", "JPN.3.0.0.0.ZCPIN", "FIN.3.0.0.0.ZCPIN", "AUT.3.0.0.0.ZCPIN", "NOR.3.0.0.0.ZCPIN", "SWE.3.0.0.0.ZCPIN", "DNK.3.0.0.0.ZCPIN", "NLD.3.0.0.0.ZCPIN", "KOR.3.0.0.0.ZCPIN", "ESP.3.0.0.0.ZCPIN", "ISL.3.0.0.0.ZCPIN", "EA21.3.0.0.0.ZCPIN")


data_cpi_ameco <- 
codes_variables %>% 
  map_dfr(~rdb(paste("AMECO/ZCPIN", .x, sep = "/"))) %>% 
  select(Country, original_period, value) %>% 
  drop_na()

data_cpi_ameco$original_period = as.numeric(data_cpi_ameco$original_period)

data_cpi_ameco <- 
  data_cpi_ameco %>% 
  group_by(Country) %>% 
  mutate(
    inflation_rate = c(NA, diff(log(value)))
  ) %>% 
  ungroup()

library(gghighlight)

data_cpi_ameco %>% 
  filter(original_period %in% c(1961:2025)) %>% 
  ggplot(aes(x = original_period, y = inflation_rate, color = Country))+
  geom_line(size = 1)+
  geom_point()+
  gghighlight(Country %in% c("Switzerland"))



```


*Pointer la resistance particuliere en cas de crise en citant rapport ocde, fmi, seco. Voir si resistance plus élevée qu'au japon*



```{r growth rate cpi}
#| echo: false
#| warning: false
#| message: false

data_cpi_ameco |>
  filter(original_period %in% c(2000:2025)) %>% 
  group_by(Country) |>
  summarise(avg_inflation = mean(inflation_rate, na.rm = TRUE), .groups = "drop") |>
  arrange(avg_inflation)

```



```{r hcpi ameco}
#| echo: false
#| warning: false
#| message: false

library(rdbnomics)
library(tidyverse)

codes_variables = c("CHE.1.0.0.0.ZCPIH", "DEU.1.0.0.0.ZCPIH", "EU27.1.0.0.0.ZCPIH", "FRA.1.0.0.0.ZCPIH", "AUT.1.0.0.0.ZCPIH")

data_hcpi_ameco <- 
codes_variables %>% 
  map_dfr(~rdb(paste("AMECO/ZCPIH", .x, sep = "/"))) %>% 
  select(Country, original_period, value) %>% 
  drop_na()

data_hcpi_ameco$original_period = as.numeric(data_hcpi_ameco$original_period)


data_hcpi_ameco %>% 
  filter(original_period %in% c(2005:2025)) %>% 
  ggplot(aes(x = original_period, y = value, color = Country, group = Country))+
  geom_line()+
  geom_point()+
  theme_minimal()


```


What accounts for Switzerland's remarkable and enduring resilience to inflationary pressures? The literature has advanced several competing hypotheses, each emphasising different structural, institutional, or macroeconomic mechanisms. The most prominent of these are examined in turn below.

## The Safe-Haven Currency Hypothesis

The Swiss franc is widely regarded as a safe-haven currency, serving as a global store of value during periods of economic and geopolitical uncertainty. Its status as a refuge currency dates back to at least the First World War, when Switzerland's comparative monetary stability stood in sharp contrast to the widespread currency upheaval that followed the Great War. Switzerland's non-belligerence contributed substantially to the franc's attractiveness, a status further consolidated by the stability of Swiss socio-political institutions. While the franc's safe-haven role was reinforced progressively throughout the post-war period, it became fully manifest with the collapse of the Bretton Woods system in 1973: the dissolution of the dollar peg triggered a sustained and substantial nominal appreciation of the Swiss franc. In theory, this secular appreciation should exert downward pressure on inflation by lowering the relative price of imported goods and services. However, the empirical literature has called this mechanism into question by documenting a strikingly incomplete exchange rate pass-through. @auer_exchange_2021 find that a 1% appreciation of the Swiss franc translates into a decline in retail import prices of only 0.55% after two quarters. @oktay_heterogeneity_2022 estimates an overall pass-through coefficient of 0.12 across the full CPI basket, implying that only 12% of an EUR/CHF exchange rate shock is transmitted to Swiss consumer prices. The pass-through is thus substantially incomplete, which considerably weakens the case for currency appreciation as a primary driver of Switzerland's low inflation performance.

## The Administered Prices Hypothesis

A second line of explanation points to the comparatively large share of administered prices within Switzerland's HICP basket. Switzerland exhibits one of the highest proportions of fully or partially regulated prices among European economies. In the wake of the supply shocks triggered by the COVID-19 pandemic and the Russian invasion of Ukraine, post-Keynesian economists — most prominently Isabella Weber — argued in favour of price controls as an instrument to combat cost-push inflation, particularly when amplified by sellers' inflation dynamics in which firms exploit crisis conditions to expand profit margins [@weber_sellers_2023]. On this reading, the prevalence of administered prices in Switzerland constitutes a structural mechanism that dampens inflationary pressures by insulating a significant portion of the consumption basket from market price volatility.

However, recent empirical evidence suggests that administered prices are not the primary driver of Switzerland's low inflation performance. The vast majority of Swiss administered prices correspond to healthcare goods and services, which are classified as non-administered in several euro area countries, or attributed to the public sector rather than private consumption on account of their public funding. Crucially, not only have administered prices risen less in Switzerland than in the euro area, but so have non-administered prices. @auer_how_2025 show that administered prices account for less than one quarter of the observed inflation differential between Switzerland and the euro area. The dominant share of the differential is therefore explained by subdued price dynamics in the non-administered segment, pointing to deeper structural factors at work.


## The CPI Construction Hypothesis

A third hypothesis holds that Switzerland's low measured inflation is in part a statistical artefact, reflecting the specific structure and expenditure weights of the Swiss CPI basket. Critics point, in particular, to the relatively low weight assigned to energy and to the exclusion of certain major expenditure items, most notably compulsory health insurance premiums, from the standard consumer price index. Some economists have addressed this by re-estimating Swiss inflation using the expenditure weights applied in other countries, such as Germany. Nevertheless, even under these counterfactual weighting schemes, Swiss inflation remains substantially lower than in comparable economies, suggesting that measurement conventions account for only a modest fraction of the observed differential.

## The Wage-Price Spiral and Sellers' Inflation Hypothesis

A fourth explanation focuses on the limited role played by domestic wage-price dynamics in Switzerland. Given the strongly export-oriented character of Swiss production, increases in unit labour costs tend to be reflected primarily in export prices rather than in domestic consumer prices. Moreover, approximately 40% of the Swiss CPI basket — notably rents and healthcare — lies outside the typical scope of wage-price spiral mechanisms. @leutert_wageprice_2026 find that, once imported inflation and inflation expectations are controlled for, the pass-through from unit labour costs to the CPI is negligible in Switzerland. These authors conclude that Swiss CPI dynamics are driven predominantly by fluctuations in import prices — themselves shaped by movements in oil prices, foreign inflation, the exchange rate, and inflation expectations — rather than by domestic wage-setting processes. This finding is broadly consistent with the sellers' inflation framework developed by @weber_sellers_2023: while the data do not support a significant spiral between wages and domestic consumer prices, the question of whether Swiss firms have exploited supply-side disruptions to expand profit margins warrants closer empirical scrutiny.




## The Energy Mix Hypothesis

Fluctuations in oil and natural gas prices rank among the most powerful drivers of consumer price inflation. As fossil fuels constitute a critical intermediate input in the production of a wide range of goods and services, energy price shocks propagate rapidly through value chains and into the broader price level. The supply disruptions triggered by the COVID-19 pandemic and, more forcefully, by the Russian invasion of Ukraine, were instrumental in the resurgence of inflation across Europe from 2021 onwards.

Switzerland's comparatively limited dependence on fossil fuels — owing to its substantial reliance on hydroelectric and nuclear power generation — implies that energy price shocks transmit less forcefully to domestic consumer prices than in more fossil-fuel-dependent economies. This structural characteristic provides a natural buffer against the kind of cost-push dynamics that drove the post-pandemic inflation episode across much of the continent.

*While this explanation is intuitively compelling, it is complicated by the observation that several countries with even greener energy mixes, Iceland, Sweden, Norway, and Finland, nonetheless experienced higher inflation rates than Switzerland. The relevant distinguishing factor appears to be institutional rather than merely technical: Switzerland operates a regulated public energy market that anchors electricity and energy retail prices to domestic production costs. By contrast, the Nordic countries, despite their predominantly renewable generation capacity, are integrated into the European internal energy market, where prices are set at the marginal cost of the most expensive dispatched source, typically imported gas or coal. Their generation is low-carbon, but the price of that energy is nonetheless aligned with fossil fuel benchmarks. Switzerland's institutional insulation from this pricing mechanism provides an additional anti-inflationary buffer that a green energy mix alone cannot replicate.*

*==> est-ce que l'on peut tester cela avec le modele de prix de leontief ?*

A further dimension of this hypothesis concerns energy intensity. Switzerland ranks as the advanced economy with the lowest energy intensity of production, consuming exceptionally little energy per unit of gross value added. Low energy intensity implies that supply-side price shocks translate less directly into consumer price inflation and propagate less extensively through domestic value chains. A recent study by the State Secretariat for Economic Affairs (SECO) confirms that this efficiency advantage is pervasive across all sectors of the Swiss economy, agriculture, industry, services, and households alike. Swiss manufacturing, for instance, consumes less than one-third of the energy used in French manufacturing and less than half of that used in German manufacturing per unit of gross value added. This structural advantage has been further reinforced since 2010 by the expansion of the pharmaceutical sector, which is characterised by relatively low energy intensity and whose share of Swiss manufacturing value added rose from 15% to 31% between 2010 and 2019.

```{r energy mix comparison}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)

df_lc <- read.csv("low-carbon-share-energy.csv")

df_lc |>
  mutate(
    Entity = fct_reorder(Entity, Low.carbon.energy),
    groupe = case_when(
      Entity == "Switzerland" ~ "Suisse",
      Entity == "Europe"      ~ "Moyenne Europe",
      TRUE                    ~ "Autres pays"
    )
  ) |>
  ggplot(aes(y = Entity, x = Low.carbon.energy, fill = groupe)) +
  geom_col(width = 0.6) +
  geom_vline(
    xintercept = df_lc$Low.carbon.energy[df_lc$Entity == "Europe"],
    linetype = "dashed", colour = "grey40", linewidth = 0.6
  ) +
  geom_text(aes(label = paste0(round(Low.carbon.energy, 1), "%")),
            hjust = -0.2, size = 3.2) +
  scale_x_continuous(limits = c(0, 92), labels = scales::label_percent(scale = 1)) +
  scale_fill_manual(values = c(
    "Suisse"         = "#c0392b",
    "Moyenne Europe" = "#7f8c8d",
    "Autres pays"    = "#2980b9"
  )) +
  labs(
    title   = "Part des énergies bas carbone dans la consommation primaire (2024)",
    x       = "% de l'énergie primaire",
    y       = NULL,
    fill    = NULL,
    caption = "Source : Our World in Data / Energy Institute"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.major.y = element_blank())
```

```{r energy mix switzerland}
#| echo: false
#| warning: false
#| message: false

library(jsonlite)

# Fetch the data
df <- read.csv("https://ourworldindata.org/grapher/share-energy-source-sub.csv?v=1&csvType=full&useColumnShortNames=true")

# Fetch the metadata
metadata <- fromJSON("https://ourworldindata.org/grapher/share-energy-source-sub.metadata.json?v=1&csvType=full&useColumnShortNames=true")


energy_colors <- c(
  "Oil"              = "#4e3b2e",
  "Coal"             = "#6b6b6b",
  "Gas"              = "#e07b39",
  "Nuclear"          = "#8e44ad",
  "Hydro"            = "#2980b9",
  "Wind"             = "#1abc9c",
  "Solar"            = "#f1c40f",
  "Other Renewables" = "#27ae60"
)

df %>%
  filter(entity == "Switzerland") %>%
  pivot_longer(
    cols = ends_with("_pct_equivalent_primary_energy"),
    names_to = "source",
    values_to = "share"
  ) %>%
  mutate(source = source %>%
           str_remove("__pct_equivalent_primary_energy") %>%
           str_replace_all("_", " ") %>%
           str_to_title()) %>%
  ggplot(aes(x = year, y = share, color = source)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = energy_colors) +
  labs(
    title = "Switzerland: Energy Mix",
    subtitle = "Share of primary energy by source",
    x = NULL, y = "% of primary energy",
    color = NULL,
    caption = "Source: Our World in Data, Energy Institute"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position   = "right",
    plot.title        = element_text(face = "bold"),
    panel.grid.minor  = element_blank()
  )

```

```{r inflation hcpi}
#| echo: false
#| warning: false
#| message: false

data2 <- 
data_hcpi_ameco %>% 
  group_by(Country) %>% 
  mutate(
    inflation_rate = c(NA,diff(log(value)))
  ) %>% 
  ungroup()

data2 %>% 
  filter(original_period %in% c(2005:2025)) %>% 
  ggplot(aes(x = original_period, y = inflation_rate, color = Country, group = Country))+
  geom_line()+
  geom_point()

```



```{r admin prices}
#| echo: false
#| warning: false
#| message: false


library(tidyverse)
hicp_admin_prices <- read_csv("hicp_admin_prices.csv") 

hicp_admin_prices %>% 
  filter(geo %in% c("Switzerland", "European Union - 27 countries (from 2020)") & coicop18 %in% c("Administered prices", "Fully administered prices", "Mainly administered prices", "Administered prices - energy and food") & TIME_PERIOD %in% c(2015:2025)) %>% 
  ggplot(aes(x = TIME_PERIOD, y = OBS_VALUE/10, color = geo, shape = coicop18))+
  geom_line()+
  geom_point()+
  facet_wrap(~geo)
```


```{r input output analysis}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)

CHE2022ttl <- read_csv("CHE2022ttl.csv")

# ── Dimensions ────────────────────────────────────────────────────────────────
n          <- 50
col_labels <- names(CHE2022ttl)[2:(n + 1)]

# ── Intermediate transaction matrix Z and total output x ─────────────────────
Z      <- CHE2022ttl[1:n, 2:(n + 1)] |> as.matrix()
x      <- CHE2022ttl[CHE2022ttl$...1 == "OUTPUT", 2:(n + 1)] |> as.numeric()
x_safe <- ifelse(x == 0, 1, x)          # B05 (coal) has zero output in CH

# ── Technical coefficients A ──────────────────────────────────────────────────
A    <- sweep(Z, 2, x_safe, "/")         # a_ij = z_ij / x_j
hfce <- CHE2022ttl[1:n, "HFCE"] |> pull() |> as.numeric()
names(hfce) <- col_labels

# ── Sector labels ─────────────────────────────────────────────────────────────
sector_names <- c(
  A01="Crop & animal production", A02="Forestry", A03="Fishing",
  B05="Coal mining", B06="Oil & gas extraction", B07="Metal ore mining",
  B08="Other mining", B09="Mining support",
  C10T12="Food, beverage & tobacco", C13T15="Textiles & apparel",
  C16="Wood products", C17_18="Paper & printing",
  C19="Petroleum products", C20="Chemicals",
  C21="Pharmaceuticals", C22="Rubber & plastics",
  C23="Non-metallic minerals", C24A="Basic iron & steel",
  C24B="Basic non-ferrous metals", C25="Fabricated metals",
  C26="Computer & electronics", C27="Electrical equipment",
  C28="Machinery n.e.c.", C29="Motor vehicles",
  C301="Building of ships & boats", C302T309="Other transport equip.",
  C31T33="Furniture & other manuf.", D="Electricity & gas",
  E="Water & waste", F="Construction",
  G="Wholesale & retail trade", H49="Land transport",
  H50="Water transport", H51="Air transport",
  H52="Warehousing & transport support", H53="Postal services",
  I="Accommodation & food", J58T60="Publishing & broadcasting",
  J61="Telecommunications", J62_63="IT & computer services",
  K="Financial services", L="Real estate",
  M="Professional & scientific services", N="Administrative services",
  O="Public admin & defence", P="Education",
  Q="Health & social work", R="Arts & entertainment",
  S="Other services", T="Household services"
)

# ── Partition: endogenous vs. exogenous sectors ───────────────────────────────
# Following Weber et al. (2024): commodity sectors (mining, energy, petroleum)
# and rent-extracting FIRE activities are treated as exogenous — their prices
# do not respond to cost shocks originating in the domestic endogenous system.
exog_sectors  <- c("B05", "B06", "B07", "B08", "B09", "C19", "D", "K", "L")
endog_sectors <- setdiff(col_labels, exog_sectors)

exog_idx  <- which(col_labels %in% exog_sectors)
endog_idx <- which(col_labels %in% endog_sectors)
n_E       <- length(endog_idx)

# Endogenous ← endogenous sub-matrix; exogenous inputs into endogenous sectors
A_EE <- A[endog_idx, endog_idx]
A_XE <- A[exog_idx,  endog_idx]

# ── Leontief inverse for the endogenous sub-system ───────────────────────────
L_EE <- solve(diag(n_E) - A_EE)

# ── HFCE weights (denominator = total HFCE across all sectors) ───────────────
total_hfce <- sum(hfce)
w_E <- hfce[endog_idx] / total_hfce
w_X <- hfce[exog_idx]  / total_hfce

# ── Price multipliers and HFCE-weighted inflation impacts ────────────────────
# Endogenous sector j: shock to v_j raises sector i's price by L_EE[j, i]
price_mult_E <- rowSums(L_EE)
wt_impact_E  <- as.vector(L_EE %*% w_E)

# Exogenous sector k: shock to p_k propagates into endogenous system via A_XE;
# consumers also pay more directly for k (w_X term)
price_mult_X <- 1 + as.vector(A_XE %*% rowSums(L_EE))
wt_impact_X  <- w_X + as.vector(A_XE %*% wt_impact_E)

# ── Combined results ──────────────────────────────────────────────────────────
results <- bind_rows(
  tibble(sector = endog_sectors, name = sector_names[endog_sectors],
         type = "Endogenous", price_mult = price_mult_E, wt_impact = wt_impact_E),
  tibble(sector = exog_sectors,  name = sector_names[exog_sectors],
         type = "Exogenous (commodity / FIRE)",
         price_mult = price_mult_X, wt_impact = wt_impact_X)
) |> arrange(desc(wt_impact))

# ── Plot ──────────────────────────────────────────────────────────────────────
threshold <- mean(results$wt_impact) + sd(results$wt_impact)

results |>
  slice_max(wt_impact, n = 25) |>
  mutate(
    label       = paste0(sector, " — ", str_trunc(name, 32, ellipsis = "…")),
    label       = fct_reorder(label, wt_impact),
    significant = wt_impact >= threshold
  ) |>
  ggplot(aes(x = wt_impact, y = label, fill = significant, alpha = type)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = threshold, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("TRUE" = "#c0392b", "FALSE" = "#aab7b8"),
    labels = c("TRUE" = "Systemically significant", "FALSE" = "Other"),
    name   = NULL
  ) +
  scale_alpha_manual(
    values = c("Endogenous" = 1, "Exogenous (commodity / FIRE)" = 0.5),
    name   = "Treatment"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Systemically Significant Prices — Switzerland (2022)",
    subtitle = "Faded bars = exogenous sectors (commodity / FIRE); dashed line = mean + 1 SD",
    x        = "HFCE-weighted price multiplier",
    y        = NULL,
    caption  = "Source: OECD ICIO CHE2022ttl. Method: partitioned Leontief price model (Weber et al. 2024)."
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position    = "top",
    panel.grid.major.y = element_blank(),
    axis.text.y        = element_text(size = 9),
    plot.title         = element_text(size = 11, face = "bold"),
    plot.caption       = element_text(colour = "grey50", size = 7)
  )
```



```{r input output analysis 2017 ofs}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(readxl)

siot_raw <- read_excel("IO_table_ofs_2017.xlsx", sheet = "siot")

# ── Dimensions ────────────────────────────────────────────────────────────────
n <- 51   # sectors in the FSO 2017 symmetric IO table

# ── Sector codes and labels (rows 6 to 56, cols 1–2) ─────────────────────────
s_codes <- siot_raw[6:(5 + n), 1] |> pull()
s_names <- siot_raw[6:(5 + n), 2] |> pull()

# ── Intermediate transaction matrix Z, output x, HFCE ────────────────────────
# Z: rows 1:n × cols 3:(n+2); x: col 79 "Total use at basic prices";
# hfce: col 67 "Total final consumption expenditure by households"
Z    <- siot_raw[6:(5 + n), 3:(n + 2)] |> as.matrix() |> apply(c(1, 2), as.numeric)
x    <- siot_raw[6:(5 + n), 79] |> pull() |> as.numeric()
hfce <- siot_raw[6:(5 + n), 67] |> pull() |> as.numeric()

# ── Technical coefficients A ──────────────────────────────────────────────────
x_safe <- ifelse(x == 0, 1, x)
A      <- sweep(Z, 2, x_safe, "/")

# ── Partition: exogenous (commodity / FIRE) vs. endogenous ───────────────────
# Following Weber et al. (2024): mining, energy, petroleum+chemicals, finance,
# real estate treated as exogenous price-setters.
exog_sectors_2017  <- c("05 - 09", "19 - 20", "35", "64", "65", "68")
endog_sectors_2017 <- setdiff(s_codes, exog_sectors_2017)

exog_idx  <- which(s_codes %in% exog_sectors_2017)
endog_idx <- which(s_codes %in% endog_sectors_2017)
n_E       <- length(endog_idx)

A_EE <- A[endog_idx, endog_idx]
A_XE <- A[exog_idx,  endog_idx]

# ── Leontief inverse (endogenous sub-system) ──────────────────────────────────
L_EE <- solve(diag(n_E) - A_EE)

# ── HFCE weights and price multipliers ───────────────────────────────────────
total_hfce   <- sum(hfce)
w_E          <- hfce[endog_idx] / total_hfce
w_X          <- hfce[exog_idx]  / total_hfce
price_mult_E <- rowSums(L_EE)
wt_impact_E  <- as.vector(L_EE %*% w_E)
price_mult_X <- 1 + as.vector(A_XE %*% rowSums(L_EE))
wt_impact_X  <- w_X + as.vector(A_XE %*% wt_impact_E)

# ── Combined results ──────────────────────────────────────────────────────────
results_2017 <- bind_rows(
  tibble(sector = endog_sectors_2017, name = s_names[endog_idx],
         type = "Endogenous", price_mult = price_mult_E, wt_impact = wt_impact_E),
  tibble(sector = exog_sectors_2017,  name = s_names[exog_idx],
         type = "Exogenous (commodity / FIRE)",
         price_mult = price_mult_X, wt_impact = wt_impact_X)
) |> arrange(desc(wt_impact))

# ── Plot ──────────────────────────────────────────────────────────────────────
threshold_2017 <- mean(results_2017$wt_impact) + sd(results_2017$wt_impact)

results_2017 |>
  slice_max(wt_impact, n = 25) |>
  mutate(
    label       = paste0(sector, " — ", str_trunc(name, 34, ellipsis = "…")),
    label       = fct_reorder(label, wt_impact),
    significant = wt_impact >= threshold_2017
  ) |>
  ggplot(aes(x = wt_impact, y = label, fill = significant, alpha = type)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = threshold_2017, linetype = "dashed",
             colour = "grey40", linewidth = 0.5) +
  scale_fill_manual(
    values = c("TRUE" = "#c0392b", "FALSE" = "#aab7b8"),
    labels = c("TRUE" = "Systemically significant", "FALSE" = "Other"),
    name   = NULL
  ) +
  scale_alpha_manual(
    values = c("Endogenous" = 1, "Exogenous (commodity / FIRE)" = 0.5),
    name   = "Treatment"
  ) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.08))) +
  labs(
    title    = "Systemically Significant Prices — Switzerland (2017)",
    subtitle = "Faded bars = exogenous sectors (commodity / FIRE); dashed line = mean + 1 SD",
    x        = "HFCE-weighted price multiplier",
    y        = NULL,
    caption  = "Source: FSO Swiss IO table 2017. Method: partitioned Leontief price model (Weber et al. 2024)."
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position    = "top",
    panel.grid.major.y = element_blank(),
    axis.text.y        = element_text(size = 9),
    plot.title         = element_text(size = 11, face = "bold"),
    plot.caption       = element_text(colour = "grey50", size = 7)
  )
```



```{r network analysis io 2017}
#| echo: false
#| warning: false
#| message: false

library(igraph)
library(ggraph)
library(tidygraph)
library(tidyverse)

# ── Short display labels ───────────────────────────────────────────────────────
short_labels <- c(
  "01"="Agriculture", "02"="Forestry", "03"="Fishing",
  "05 - 09"="Mining", "10 - 12"="Food & tobacco",
  "13 - 15"="Textiles", "16"="Wood", "17"="Paper",
  "18"="Printing", "19 - 20"="Petrol. & chem.",
  "21"="Pharma", "22"="Rubber & plastics",
  "23"="Non-met. minerals", "24"="Basic metals",
  "25"="Fab. metals", "26"="Computer & electr.",
  "27"="Electrical equip.", "28"="Machinery",
  "29"="Motor vehicles", "30"="Other transport eq.",
  "31"="Furniture", "32"="Other manuf.",
  "33"="Repair & install.", "35"="Electricity & gas",
  "36 - 39"="Water & waste", "41 - 43"="Construction",
  "45"="Motor trade", "46"="Wholesale",
  "47"="Retail", "49 - 51"="Transport",
  "52"="Warehousing", "53"="Postal",
  "55"="Accommodation", "56"="Food service",
  "58 - 60"="Publishing & broadcast.", "61"="Telecom",
  "62 - 63"="IT services", "64"="Finance",
  "65"="Insurance", "68"="Real estate",
  "69 - 71"="Legal & engineering", "72"="R&D",
  "73 - 75"="Advertising & prof.", "77 - 82"="Admin. services",
  "84"="Public admin.", "85"="Education",
  "86"="Health", "87 - 88"="Residential care",
  "90 - 93"="Arts & recreation", "94 - 96"="Other services",
  "97 - 98"="Household services"
)

# ── Build directed network (supplier → buyer) from A matrix ───────────────────
# Edge i → j: sector i provides inputs to sector j
# Keep edges where A_ij ≥ 1% (sector i supplies ≥ 1% of j's total input)
edge_df_net <- which(A >= 0.01 & row(A) != col(A), arr.ind = TRUE) |>
  as.data.frame() |>
  dplyr::rename(from_i = row, to_j = col) |>
  dplyr::mutate(
    from   = s_codes[from_i],
    to     = s_codes[to_j],
    weight = Z[cbind(from_i, to_j)],
    a_coef = A[cbind(from_i, to_j)]
  )

node_df_net <- data.frame(
  name  = s_codes,
  label = unname(short_labels[s_codes]),
  exog  = s_codes %in% exog_sectors_2017
)

g_net <- igraph::graph_from_data_frame(
  d        = edge_df_net[, c("from", "to", "weight", "a_coef")],
  vertices = node_df_net,
  directed = TRUE
)

# ── Centrality measures ───────────────────────────────────────────────────────
igraph::V(g_net)$pagerank     <- igraph::page_rank(g_net, weights = igraph::E(g_net)$weight)$vector
igraph::V(g_net)$in_strength  <- igraph::strength(g_net, mode = "in",  weights = igraph::E(g_net)$weight)
igraph::V(g_net)$out_strength <- igraph::strength(g_net, mode = "out", weights = igraph::E(g_net)$weight)
igraph::V(g_net)$betweenness  <- igraph::betweenness(g_net, weights = 1 / igraph::E(g_net)$a_coef,
                                                      normalized = TRUE)

centrality_df_net <- igraph::as_data_frame(g_net, what = "vertices") |>
  dplyr::rename(sector = name) |>
  dplyr::mutate(
    short = short_labels[sector],
    type  = ifelse(exog, "Exogenous", "Endogenous")
  )

# ── Plot 1: Multi-measure centrality bar chart ────────────────────────────────
top_net <- centrality_df_net |>
  dplyr::filter(
    dplyr::percent_rank(pagerank)     > 0.70 |
    dplyr::percent_rank(betweenness)  > 0.70 |
    dplyr::percent_rank(in_strength)  > 0.70 |
    dplyr::percent_rank(out_strength) > 0.70
  ) |>
  dplyr::pull(sector)

centrality_df_net |>
  dplyr::filter(sector %in% top_net) |>
  dplyr::select(sector, short, type, pagerank, in_strength, out_strength, betweenness) |>
  dplyr::mutate(dplyr::across(pagerank:betweenness,
                              \(x) (x - min(x)) / (max(x) - min(x)))) |>
  tidyr::pivot_longer(pagerank:betweenness, names_to = "metric", values_to = "score") |>
  dplyr::mutate(
    metric = dplyr::recode(metric,
      pagerank     = "PageRank",
      in_strength  = "In-strength\n(backward linkages)",
      out_strength = "Out-strength\n(forward linkages)",
      betweenness  = "Betweenness"
    ),
    label = paste0(sector, " — ", short),
    label = forcats::fct_reorder(label, score, .fun = mean)
  ) |>
  ggplot(aes(x = score, y = label, fill = type)) +
  geom_col(width = 0.7) +
  facet_wrap(~metric, nrow = 1, scales = "free_x") +
  scale_fill_manual(
    values = c("Endogenous" = "#1a6faf", "Exogenous" = "#c0392b"),
    name = NULL
  ) +
  scale_x_continuous(labels = NULL, expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Supply-chain network centrality by measure — Switzerland 2017",
    subtitle = "Each metric normalised to [0,1] | Edge threshold: A ≥ 1%",
    x        = "Normalised score →",
    y        = NULL,
    caption  = "Source: FSO 2017 IO table. Direction: supplier → buyer (Z matrix, MCHF weights)."
  ) +
  theme_minimal(base_size = 9) +
  theme(
    legend.position    = "top",
    panel.grid.major.y = element_blank(),
    strip.text         = element_text(face = "bold", size = 8.5),
    plot.title         = element_text(face = "bold", size = 10.5),
    plot.caption       = element_text(colour = "grey50", size = 7),
    axis.text.y        = element_text(size = 8)
  )

# ── Plot 2: Network graph ─────────────────────────────────────────────────────
sector_group <- dplyr::case_when(
  s_codes %in% c("01","02","03","10 - 12") ~ "Agri. & food",
  s_codes %in% c("05 - 09","19 - 20","35","36 - 39") ~ "Energy & resources",
  s_codes %in% c("13 - 15","16","17","18","21","22","23","24","25",
                  "26","27","28","29","30","31","32","33") ~ "Manufacturing",
  s_codes %in% c("41 - 43") ~ "Construction",
  s_codes %in% c("45","46","47","49 - 51","52","53") ~ "Trade & transport",
  s_codes %in% c("55","56","58 - 60","61","62 - 63",
                  "64","65","68","69 - 71","72","73 - 75","77 - 82") ~ "Services (private)",
  TRUE ~ "Public & other services"
)

tg_net <- as_tbl_graph(g_net) |>
  tidygraph::activate(nodes) |>
  dplyr::mutate(
    group     = sector_group,
    pr        = igraph::page_rank(g_net, weights = igraph::E(g_net)$weight)$vector,
    short_lbl = short_labels[name]
  ) |>
  tidygraph::activate(edges) |>
  dplyr::filter(a_coef >= 0.04)   # show only stronger edges for clarity

group_colors <- c(
  "Agri. & food"            = "#2ecc71",
  "Energy & resources"      = "#e67e22",
  "Manufacturing"           = "#3498db",
  "Construction"            = "#8e44ad",
  "Trade & transport"       = "#1abc9c",
  "Services (private)"      = "#c0392b",
  "Public & other services" = "#7f8c8d"
)

set.seed(4201)
ggraph(tg_net, layout = "stress") +
  geom_edge_arc(aes(width = a_coef, alpha = a_coef),
                arrow = arrow(length = unit(2, "mm"), type = "closed"),
                end_cap = circle(3, "mm"),
                color = "grey60", strength = 0.15) +
  geom_node_point(aes(size = pr, fill = group), shape = 21,
                  color = "white", stroke = 0.4) +
  geom_node_text(aes(label = ifelse(pr > quantile(pr, 0.72), short_lbl, "")),
                 size = 2.4, repel = TRUE, max.overlaps = 20,
                 color = "grey15", fontface = "bold") +
  scale_edge_width(range = c(0.3, 2.5), guide = "none") +
  scale_edge_alpha(range = c(0.25, 0.8), guide = "none") +
  scale_size(range = c(2, 12), name = "PageRank", guide = "none") +
  scale_fill_manual(values = group_colors, name = NULL) +
  labs(
    title    = "Swiss inter-industry supply-chain network (2017)",
    subtitle = "Node size = PageRank | Edges shown: A ≥ 4% | Labels: top 28% by PageRank",
    caption  = "Source: FSO Swiss IO table 2017. Direction: supplier → buyer."
  ) +
  theme_void(base_size = 10) +
  theme(
    plot.title      = element_text(face = "bold", size = 11, hjust = 0.5),
    plot.subtitle   = element_text(color = "grey50", size = 8, hjust = 0.5,
                                   margin = margin(b = 4)),
    plot.caption    = element_text(color = "grey50", size = 7, hjust = 0.5),
    legend.position = "right",
    plot.margin     = margin(10, 10, 10, 10),
    plot.background = element_rect(fill = "white", color = NA)
  )
```



## References
