---
title: "Political Compass: what lies behind political science's most famous graphs"
author: "Celâl Güney"
format: html
editor: visual
execute: 
  freeze: true
date: "21 July 2023"
categories: [Political Science, Statistics]
draft: false
bibliography: pol_compass.bib
---

```{r}
#| echo: false
#| warning: false
#| message: false
rm(list = ls())
library(haven)
library(sjlabelled)
library(tidyverse)
library(viridis)
library(stargazer)
library(gtsummary)
library(gt)
library(brms)
library(ggrepel)
library(cowplot)
library(ggpubr)
library(lme4)
library(prediction)
library(marginaleffects)
library(ggeffects)
library(ordinal)
library(broom)
library(knitr)
library(readr)
library(gtExtras)
library(broom.mixed)
library(performance)
library(GGally)
library(questionr)
library(car)
library(effects)
library(polcompr)
library(stargazer)
library(corrplot)
library(ggrepel)

theme_set(theme_bw())
```

People who get lost or waste their time on internet have surely seen the now famous "political compass" graphs, which have now become a popular meme and perhaps the most popular one related to politics. The irony and paradox of political compass, and that is what makes it intriguing, are that it is actually and important representation in political science. Very shortly, a political compass is a mapping of political leaning and preferences based on two axes: (1) an economic axis representing the position of an individual, party, association or group on global economic issues (pro and anti redistribution, for economic planning or free market...) and (2) a "cultural" or "post-material" (a lot of different names are given to this axis) axis representing position on cultural issues, such as: "in favor or against immigration", "in favor or against gay marriage/lgbt rights/traditional values..."

Whereas most of people who use political compass do so to make memes and amuse their friends, political scientists are still seriously attached to their favorite toy and one often sees such mapping on political science articles. For instance, Oesch and Rennwald provide such graphics to map individuals political preferences for Switzerland and other countries using European Social Survey data [@oesch_electoral_2018, 795]. Pascal Sciarini, in a manual on Swiss politics, also provides such a graph using Swiss Elections Studies data [@sciarini_politique_2023, 341]. The idea that political preferences can be separated between an economic and cultural cleavage comes back at least to Inglehart [-@inglehart_silent_1971] who explained the emerging salience of cultural issues over the traditional economic divide as a consequence of post-war tremendous economic growth which made the West so wealthy that people would get rid off of any material insecurity, hence shifting the focus on cultural matters rather than economic ones.
But how do political scientists construct such graphs? The goal of this post is to explore how the economic and cultural axes are constructed, what are the data and variables used, and what kind of statistical manipulation is done with those data.


## Constructing a Political Compass

Political compass is typically constructed from ordinal variables related to economic and cultural issues and the individual self-placement of survey respondents to these issues. For instance, Oesch and Rennwald [-@oesch_electoral_2018] construct the preferences of respondents on the economic axis from an ordinal 5-points scale variable about preferences on redistribution (1 = strongly agree to reduce differences in income levels, 5 = strongly disagree). For the cultural cleavage, they took three variables: (1) the impact of immigrants on cultural life, (2) European integration and (3) freedom of gay people. One can directly see the problem with the construction of such a "cultural" axis: are here the European integration and immigration issues purely cultural issues? In my opinion, European integration cannot be considered only as cultural "openness vs closeness values" issue and should not have been included to construct the cultural axis. However, the next step after choosing the variables is to standardize them: each observed value of the dataset is subtracted by the mean of the variable and divided by its standard deviation. If more than one variable are taken into account, the variables are standardized and the average values on these three standardized variables are computed to create the final index.

To work with an example, I will work here with a dataset which is not used by the literature cited above: the Voxit studies. The latter offer a cumulative dataset on Swiss referendum voting from 1981 to 2016. The dataset contains a list of "issue positions" variables:

- In favor of state intervention or free markets
- In favor of or against income equality 
- In favor of an open or closed Switzerland 
- In favor of active equality policy between men and women 
- In favor of equal chances between Swiss people and foreigners
- In favor of environnment protection or economic growth
- In favor of a strong or weak Swiss army

All these variables are ordinal on a 7-points scale. Another variable, "p02" collected political party support (which party the respondent feels the most closed to). I will use this variable as a group to plot Swiss political parties on the economic and cultural axes depending on the average value of their supporters on the standardized economic and cultural issues variables. To construct the economic axis, I will use the "state vs market" variable. "Open vs closed Switzerland" will be used for the cultural axis.

```{r}
#| echo: false
#| warning: false
#| message: false
vox <- read_sav("F:/mapec/Master THESIS/projectB_politicalecon/data/vox/vox_cumulative_dataset/688_VoxIt_Data_CumulatedFile_Scrutins_15_121_F.sav")
```

```{r create income variable}
#| warning: false
#| message: false
#| echo: false

# Two ways to create the income variable
vox <- vox %>% 
  mutate(income = if_else(revenu == 1 & is.na(nivmena), 4,
                          if_else(revenu == 2 & is.na(nivmena), 3,
                                  if_else(revenu %in% c(3,4) & is.na(nivmena), 2,
                                          if_else(revenu == 5 & is.na(nivmena), 1,
                                                  if_else(is.na(revenu) & annee <= 1993, nivmena, NA))))))

         
vox$income <- ifelse(vox$income == 4, -1,
                    ifelse(vox$income == 3, 0,
                        ifelse(vox$income == 2, 1,
                          ifelse(vox$income == 1, 2, NA)))) # centering around zero, this will make the intercept more interpretable 


# for education, rescale the values so that the lowest level is equal to zero
vox <- vox %>% 
  mutate(education = case_when(
    educ == 1 ~ 0,
    educ == 2 ~ 1,
    educ == 3 ~ 2,
    educ == 4 ~ 3,
    educ == 5 ~ 4,
    educ == 6 ~ 5,
    is.na(educ) ~ NA
  ))

```

```{r party dummies and other recoding}
#| warning: false
#| message: false
#| echo: false


missing_values <- function(data, variable, group) { 
  
   table <- data %>% 
     group_by({{group}}) %>% 
     count({{variable}}) %>% 
     mutate(prop = n/sum(n)) %>% 
     filter(is.na({{variable}})) %>% 
     ungroup()
  
   plot <- 
     ggplot(data = table, aes(x = {{group}}, y = prop))+
     geom_col()+
     geom_label(aes(x = {{group}}, y = prop, label = {{group}}))
   print(plot)
  
}

```


Here are all the steps to transform the variable, compute the average values for each year for each political party support and how to plot the final political compass.
```{r political compass}
#| warning: false
#| message: false


strdz <- function(x){
  (x - mean(x, na.rm = TRUE))/(sd(x, na.rm = TRUE))
} # create a function which transfrom x into standardize variable by subtracting the mean and dividing by standard deviation


vox <- 
vox %>% 
  group_by(annee) %>% # group by year
  mutate(state_market = strdz(val11), # compute the standardization transformation and store into new variable
         open_closed = strdz(val14),
         envi_vs_growth = strdz(val10),
         inegalite_hf = strdz(val16),
         weak_army = strdz(val3),
         inc_equality = strdz(val5),
         no_chance_foreign = strdz(val9)) %>% 
  ungroup()

#### create a table: each row (observation) is a political party repeated for each year in which data were available (1996 to 2016)
values <- 
vox %>% 
  group_by(annee, as_character(p02)) %>% # average computed across both year and parties
  summarise(state_market = mean(state_market, na.rm = TRUE), # compute the average of standardized state vs market variable in each year for each party support group.
            open_closed = mean(open_closed, na.rm = TRUE),
            envi_vs_growth = mean(envi_vs_growth, na.rm = TRUE),
            inegalite_hf = mean(inegalite_hf, na.rm = TRUE),
            weak_army = mean(weak_army, na.rm = TRUE),
            inc_equality = mean(inc_equality, na.rm = TRUE),
            no_chance_foreign = mean(no_chance_foreign, na.rm = TRUE)) %>% 
  ungroup() %>% 
  drop_na() %>% 
  rename("party" = `as_character(p02)`)



```

Before plotting the political compass, it could be interesting to see the correlation between the average values through the years. If two variables are positively correlated, that means there are chances that, given any year, when one has high value (average of the standardized issue position variable for each respondent having declared to support the Swiss's People party for instance) the other will have a high value as well:

```{r corrplot}
#| echo: false
#| eval: false

values %>% 
  dplyr::select(-annee, -party) %>% 
  cor() %>% 
  corrplot(order = "hclust")
```
The correlation plots give coherent results overall. For example, the average of party supporters who declared not being in favor of equal chance for foreigners are also supportive of a isolated Switzerland (see "no_chance-foreign" - "open_closed").


```{r}
#| echo: false
values$party[values$party == "PLR/FDP Die Liberalen"] <- "PLR"
```


Let's now plot a political compass for a given year. I choose the first year available: 1996. I also filter political parties to select only the most important ones: the Swiss Social Democratic Party (PSS/SPS), the Radicals (PRD), The Swiss Peoples Party (UDC/SVP), the Liberals, The Catholic-Conservatives (PDC), the Greens (PES/GPS), the Evangelical party (PEP/EVP).

```{r}
#| warning: false
#| message: false

values %>% 
  filter(annee %in% c(1996) & party %in% c("UDC/SVP", "PDC/CVP", "PLS/LPS", "PSS/SPS", "PRD/FDP", "PES/GPS", "PEP/EVP")) %>% 
  ggplot()+
  aes(x = state_market, y = open_closed, label = party)+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 0)+
  geom_label()+
  labs(x = "State intervention - Free market",
       y = "Open vs Closed Switzerland",
       title = "Swiss Political Compass",
       subtitle = "1996",
       caption = "Data source: Voxit")+
    xlim(-1,1)+
    ylim(-1,1)


```

```{r pol compass2}
#| echo: false
#| eval: false

# other possible way to compute?
values2 <- 
vox %>% 
  group_by(annee, as_character(p02)) %>% 
  reframe(val11_scaled = mean(val11, na.rm = TRUE),
          val14_scaled = mean(val14, na.rm = TRUE)) %>% 
  ungroup() %>% 
  rename("party" = `as_character(p02)`) %>% 
  mutate(z_eco = strdz(val11_scaled),
         z_values = strdz(val14_scaled))
```

```{r}
#| eval: false
#| echo: false


vote_eee <- 
vox %>% 
  filter(annee == 1992 & mois == 12 & jour == 6) %>% 
  mutate(vote_yes = case_when(
    a02x1 == 1 ~ 1,
    a02x1 %in% c(2, 3) ~ 0,
    is.na(a02x1) ~ NA
  ),
  vote_yes2 = case_when(
    decx1 == 1 ~ 1,
    decx1 %in% c(2, 3) ~ 0,
    is.na(decx1) ~ NA
  ),
  urb = case_when(
    vilcamp == 1 ~ 1,
    vilcamp == 2 ~ 0
  ),
  nivmena2 = case_when(
    nivmena == 4 ~ 0,
    nivmena == 3 ~ 1,
    nivmena == 2 ~ 2,
    nivmena == 1 ~ 3,
    is.na(nivmena) ~ NA
  ),
  educ2 = case_when(
    educ == 1 ~ 0,
    educ == 2 ~ 1,
    educ == 3 ~ 2,
    educ == 6 ~ 3,
  ),
  lr_pos = p04 - median(vote_eee$p04, na.rm = TRUE))
```
