---
title: "Input output analysis"
author: "Celâl Güney"
format: html
editor: visual
draft: true
---

```{r}
library(tidyverse)
options(scipen = 999)
setwd("~/GitHub/celalguney/posts/input output analysis")

CHE2022ttl <- read_csv("CHE2022ttl.csv") %>% column_to_rownames(var = "...1")
```


```{r compute technical coefficient}

Z <- as.matrix(CHE2022ttl[1:50, 1:50])       
x <- as.numeric(CHE2022ttl["OUTPUT", 1:50])   

technical_coef <- sweep(Z, 2, x, "/")        

technical_coef[is.nan(technical_coef)] <- 0

```

## Leontief Inverse

The Leontief inverse $$L = (I - A)^{-1}$$ captures both direct and indirect production requirements. An element $$l_{ij}$$ gives the total output required from sector $$i$$ to satisfy one unit of final demand for sector $$j$$.

```{r leontief inverse}

n <- nrow(technical_coef)
leontief_inverse <- solve(diag(n) - technical_coef)

```

## Leontief Price Model

The demand-pull model simulates quantity effects. To capture how an external oil price shock propagates as **cost increases** across sectors, we use the Leontief price model:

$$\Delta p = L^T \cdot \Delta v$$

where $$\Delta v_j = \sum_{k \in \text{oil}} a_{kj} \cdot \Delta p^{ext}_k$$ is the direct cost increase per unit of output in sector $$j$$ due to the oil price shock, and $$L^T$$ is the transpose of the domestic Leontief inverse.

We use the OECD official domestic Leontief inverse to ensure correctness.

```{r price model}

# Load OECD domestic Leontief inverse
CHE2022LEONTFD <- read_csv("CHE2022LEONTFD.csv")

leontief_dom <- CHE2022LEONTFD |>
  tibble::column_to_rownames("...1") |>
  as.matrix()

leontief_dom_50 <- leontief_dom[1:50, ]   # drop TOTAL row

# Derive A_dom from the OECD domestic Leontief: A = I - L^{-1}
A_dom <- diag(50) - solve(leontief_dom_50)

# Oil/energy sectors: B06 (crude extraction), C19 (petroleum refining), D (energy supply)
oil_sectors <- c("B06", "C19", "D")
oil_idx     <- which(colnames(leontief_dom_50) %in% oil_sectors)

# Direct cost shock per unit of output in each sector j
# = oil input share × 20% external price increase
shock_pct <- 0.20
delta_v   <- colSums(A_dom[oil_idx, , drop = FALSE]) * shock_pct

# Price model propagation: Δp = L^T · Δv
delta_p <- t(leontief_dom_50) %*% delta_v

price_shock_result <- data.frame(
  sector  = colnames(leontief_dom_50),
  delta_p = as.numeric(delta_p) * 100   # in percent
) |>
  arrange(desc(delta_p))

ggplot(price_shock_result, aes(x = delta_p, y = reorder(sector, delta_p))) +
  geom_col() +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  labs(
    title    = "Price Impact of a +20% External Oil Price Shock",
    subtitle = "Leontief price model \u2014 OECD domestic Leontief inverse",
    x        = "Estimated price increase (%)",
    y        = NULL
  )

```
